---
title: "Water Point Functionality Survey"
subtitle: "Comprehensive Assessment of Water Infrastructure Performance"
author: "Water Point Survey Team"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Project Overview

This website presents a comprehensive analysis of water point functionality data collected through the **mWater platform**. The survey was conducted to assess the operational status, accessibility, and service provision of water infrastructure across surveyed communities.

### Key Objectives

- **Assess water point functionality** across different geographic locations
- **Evaluate service provider effectiveness** and maintenance systems
- **Identify common problems** affecting water point operations
- **Analyze water quality issues** and their prevalence
- **Support evidence-based decision making** for water infrastructure investments

## Data Collection Platform

The data was collected using **mWater**, a comprehensive water monitoring platform that enables:

- **Real-time data collection** through mobile devices
- **GPS coordinate capture** for precise location mapping
- **Photo documentation** of water point conditions
- **Standardized survey protocols** ensuring data quality
- **Cloud-based data storage** and management

## Dataset Summary

```{r}
#| echo: false
#| message: false
#| warning: false

# Load required libraries
library(readr)
library(dplyr)
library(knitr)
library(DT)

# Read the processed data
data <- read_csv("data/wpf_processed.csv")

# Create summary statistics
summary_stats <- data.frame(
  Metric = c("Total Water Points", "Survey Period", "Geographic Coverage", "Data Collection Method", "Quality Assurance"),
  Value = c(
    paste(nrow(data), "water points"),
    "April 2022",
    "Regional survey area",
    "mWater mobile platform",
    "Multi-stage validation process"
  )
)

kable(summary_stats, caption = "Dataset Overview")
```

## Key Findings at a Glance

### Functional Status Distribution

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)

# Functional status distribution
func_status <- data %>%
  count(functional_status) %>%
  mutate(percentage = round(n/sum(n)*100, 1))

ggplot(func_status, aes(x = reorder(functional_status, n), y = n, fill = functional_status)) +
  geom_col() +
  geom_text(aes(label = paste0(n, "\n(", percentage, "%)")), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Water Point Functional Status Distribution",
       x = "Functional Status",
       y = "Number of Water Points") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_viridis_d()
```

### Geographic Distribution

```{r}
#| echo: false
#| message: false
#| warning: false

library(leaflet)
library(plotly)

# Convert coordinates to numeric
data <- data %>%
  mutate(
    lat = as.numeric(latitude),
    lon = as.numeric(longitude)
  ) %>%
  filter(!is.na(lat), !is.na(lon))

# Create interactive map
map <- leaflet(data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon, 
    lat = ~lat,
    color = ~case_when(
      functional_status == "FUNC" ~ "green",
      functional_status == "PARTIAL" ~ "orange", 
      functional_status == "NOT_FUNC" ~ "red",
      TRUE ~ "gray"
    ),
    radius = 5,
    fillOpacity = 0.8,
    popup = ~paste0(
      "<b>Water Point ID:</b> ", water_point_id, "<br>",
      "<b>Status:</b> ", functional_status, "<br>",
      "<b>Households:</b> ", households_count, "<br>",
      "<b>Problem:</b> ", current_problem
    )
  ) %>%
  addLegend(
    "bottomright",
    colors = c("green", "orange", "red", "gray"),
    labels = c("Functional", "Partial", "Not Functional", "Other"),
    title = "Water Point Status"
  )

map
```

### Service Provider Distribution

```{r}
#| echo: false
#| message: false
#| warning: false

# Service provider analysis
service_data <- data %>%
  filter(!is.na(service_provider_type), service_provider_type != "NA") %>%
  count(service_provider_type, sort = TRUE) %>%
  head(8)

p1 <- ggplot(service_data, aes(x = reorder(service_provider_type, n), y = n, fill = service_provider_type)) +
  geom_col() +
  coord_flip() +
  labs(title = "Water Point Service Providers",
       x = "Service Provider Type",
       y = "Number of Water Points") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_viridis_d()

ggplotly(p1)
```

### Problems by Functionality Status

```{r}
#| echo: false
#| message: false
#| warning: false

# Problem analysis
problem_data <- data %>%
  filter(!is.na(current_problem), current_problem != "NA") %>%
  count(functional_status, current_problem) %>%
  group_by(functional_status) %>%
  slice_max(n, n = 3) %>%
  ungroup()

p2 <- ggplot(problem_data, aes(x = reorder(current_problem, n), y = n, fill = functional_status)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~functional_status, scales = "free") +
  labs(title = "Most Common Problems by Functional Status",
       x = "Problem Type",
       y = "Number of Water Points") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_viridis_d()

ggplotly(p2)
```

## Data Quality Enhancements

This dataset has undergone comprehensive quality improvements:

### ✅ **Standardization**
- Functional status coded to standard categories (FUNC, PARTIAL, NOT_FUNC, ABANDONED)
- Boolean fields standardized to consistent Yes/No format
- Problem categories grouped into logical taxonomies

### ✅ **Validation**
- GPS coordinates validated and outliers removed
- Household counts validated within reasonable ranges (1-500)
- Cross-field logical consistency checks implemented

### ✅ **Enhancement**
- Multi-value service provider fields expanded to boolean indicators
- Missing data consistently marked as "NA"
- Date formats standardized to DD/MM/YYYY

### ✅ **Documentation**
- Comprehensive audit trail of all changes
- Data completeness analysis conducted
- Processing scripts provided for reproducibility

## Key Variables

The dataset includes the following primary variables:

- **`water_point_id`**: Unique identifier for each water point
- **`functional_status`**: Current operational status (FUNC/PARTIAL/NOT_FUNC/ABANDONED)
- **`latitude/longitude`**: GPS coordinates for mapping
- **`current_problem`**: Categorized problems (EQUIPMENT_FAILURE, LOW_FLOW, STRUCTURAL_ISSUE, etc.)
- **`households_count`**: Number of households served
- **`service_provider_type`**: Type of maintenance/operation provider
- **`quality_issues_type`**: Water quality concerns (ODOR, SALINITY, TURBIDITY, etc.)
- **`submitted_date`**: Survey completion date

## Navigation

Use the navigation bar to explore different sections:

- **[Data Overview](data-overview.qmd)**: Detailed dataset statistics and distributions
- **[Analysis](analysis.qmd)**: In-depth analysis and visualizations
- **[Methodology](methodology.qmd)**: Survey methodology and data collection process

## About mWater

[mWater](https://www.mwater.co/) is a comprehensive water monitoring platform that enables organizations to:

- Collect standardized water point data
- Monitor water quality and infrastructure
- Generate real-time reports and dashboards
- Support evidence-based water sector planning
- Facilitate collaboration between stakeholders

This survey demonstrates the platform's capability to generate high-quality, actionable water infrastructure data for decision-making and planning purposes.

---

*Last updated: `r format(Sys.Date(), "%B %d, %Y")`*