---
title: "Water Point Functionality Survey"
subtitle: "Comprehensive Assessment of Water Infrastructure Performance"
author: "Water Point Survey Team"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Project Overview

This website presents a comprehensive analysis of water point functionality data collected through the **mWater platform**. The survey was conducted to assess the operational status, accessibility, and service provision of water infrastructure across surveyed communities.

### Key Objectives

- **Assess water point functionality** across different geographic locations
- **Evaluate service provider effectiveness** and maintenance systems
- **Identify common problems** affecting water point operations
- **Analyze water quality issues** and their prevalence
- **Support evidence-based decision making** for water infrastructure investments

## Data Collection Platform

The data was collected using **mWater**, a comprehensive water monitoring platform that enables:

- **Real-time data collection** through mobile devices
- **GPS coordinate capture** for precise location mapping
- **Photo documentation** of water point conditions
- **Standardized survey protocols** ensuring data quality
- **Cloud-based data storage** and management

## Dataset Summary

```{r}
#| echo: false
#| message: false
#| warning: false

# Load required libraries
library(readr)
library(dplyr)
library(knitr)
library(DT)

# Read the processed data
data <- read_csv("wpf_processed.csv")

# Create summary statistics
summary_stats <- data.frame(
  Metric = c("Total Water Points", "Survey Period", "Geographic Coverage", "Data Collection Method", "Quality Assurance"),
  Value = c(
    paste(nrow(data), "water points"),
    "April 2022",
    "Regional survey area",
    "mWater mobile platform",
    "Multi-stage validation process"
  )
)

kable(summary_stats, caption = "Dataset Overview")
```

## Key Findings at a Glance

### Functional Status Distribution

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)

# Functional status distribution
func_status <- data %>%
  count(functional_status) %>%
  mutate(percentage = round(n/sum(n)*100, 1))

ggplot(func_status, aes(x = reorder(functional_status, n), y = n, fill = functional_status)) +
  geom_col() +
  geom_text(aes(label = paste0(n, "\n(", percentage, "%)")), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Water Point Functional Status Distribution",
       x = "Functional Status",
       y = "Number of Water Points") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_viridis_d()
```

### Geographic Distribution

```{r}
#| echo: false
#| message: false
#| warning: false

# Geographic summary
geo_summary <- data %>%
  summarise(
    min_lat = min(as.numeric(latitude), na.rm = TRUE),
    max_lat = max(as.numeric(latitude), na.rm = TRUE),
    min_lon = min(as.numeric(longitude), na.rm = TRUE),
    max_lon = max(as.numeric(longitude), na.rm = TRUE),
    coverage_area = paste0(
      "Latitude: ", round(min_lat, 2), "° to ", round(max_lat, 2), "°\n",
      "Longitude: ", round(min_lon, 2), "° to ", round(max_lon, 2), "°"
    )
  )

cat("**Survey Coverage Area:**\n")
cat(geo_summary$coverage_area)
```

## Data Quality Enhancements

This dataset has undergone comprehensive quality improvements:

### ✅ **Standardization**
- Functional status coded to standard categories (FUNC, PARTIAL, NOT_FUNC, ABANDONED)
- Boolean fields standardized to consistent Yes/No format
- Problem categories grouped into logical taxonomies

### ✅ **Validation**
- GPS coordinates validated and outliers removed
- Household counts validated within reasonable ranges (1-500)
- Cross-field logical consistency checks implemented

### ✅ **Enhancement**
- Multi-value service provider fields expanded to boolean indicators
- Missing data consistently marked as "NA"
- Date formats standardized to DD/MM/YYYY

### ✅ **Documentation**
- Comprehensive audit trail of all changes
- Data completeness analysis conducted
- Processing scripts provided for reproducibility

## Key Variables

The dataset includes the following primary variables:

- **`water_point_id`**: Unique identifier for each water point
- **`functional_status`**: Current operational status (FUNC/PARTIAL/NOT_FUNC/ABANDONED)
- **`latitude/longitude`**: GPS coordinates for mapping
- **`current_problem`**: Categorized problems (EQUIPMENT_FAILURE, LOW_FLOW, STRUCTURAL_ISSUE, etc.)
- **`households_count`**: Number of households served
- **`service_provider_type`**: Type of maintenance/operation provider
- **`quality_issues_type`**: Water quality concerns (ODOR, SALINITY, TURBIDITY, etc.)
- **`submitted_date`**: Survey completion date

## Navigation

Use the navigation bar to explore different sections:

- **[Data Overview](data-overview.qmd)**: Detailed dataset statistics and distributions
- **[Analysis](analysis.qmd)**: In-depth analysis and visualizations
- **[Methodology](methodology.qmd)**: Survey methodology and data collection process

## About mWater

[mWater](https://www.mwater.co/) is a comprehensive water monitoring platform that enables organizations to:

- Collect standardized water point data
- Monitor water quality and infrastructure
- Generate real-time reports and dashboards
- Support evidence-based water sector planning
- Facilitate collaboration between stakeholders

This survey demonstrates the platform's capability to generate high-quality, actionable water infrastructure data for decision-making and planning purposes.

---

*Last updated: `r format(Sys.Date(), "%B %d, %Y")`*